DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-02-21 manuscript/manuscript.tex
format titles and subtitles and add edits to cross validation section

Word count: 5956

@@

\hypertarget{searching-a-chronogram-database}{%
\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}

At the time of writing of this manuscript
(Feb 20, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

\hypertarget{choosing-a-topology}{%
\subsection{Choosing a topology}\label{choosing-a-topology}}

@@

\hypertarget{applying-secondary-calibrations}{%
\subsection{Applying secondary calibrations}\label{applying-secondary-calibrations}}

@@

\hypertarget{dating-a-tree-topology}{%
\subsection{Dating a tree topology}\label{dating-a-tree-topology}}

\hypertarget{with-no-branch-lengths}{%
\subsubsection{With no branch lengths}\label{with-no-branch-lengths}}

When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008), which has proven useful for ecological analyses that require age data (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.

@@

\hypertarget{dating-a-tree-with-branch-lengths}{%
\subsubsection{Dating a tree with branch lengths}\label{dating-a-tree-with-branch-lengths}}

Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain markers for input taxa.
By default, mined genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015). Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older nodes (Forest et al., 2005). The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
The BOLD workflow is currently only available on DateLife's R package and not on the web application.

@@

\hypertarget{visualizing-results}{%
\subsection{Visualizing results}\label{visualizing-results}}

@@

\hypertarget{a-small-example}{%
\subsection{A small example}\label{a-small-example}}

\hypertarget{creating-a-search-query-1}{%
\subsubsection{Creating a search query}\label{creating-a-search-query-1}}

We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.

\hypertarget{searching-the-database}{%
\subsubsection{Searching the database}\label{searching-the-database}}

DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

\hypertarget{summarizing-search-results-1}{%
\subsubsection{Summarizing search results}\label{summarizing-search-results-1}}

DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).

@@

\hypertarget{an-example-with-the-family-of-true-finches}{%
\subsection{An example with the family of true finches}\label{an-example-with-the-family-of-true-finches}}

\hypertarget{creating-a-query}{%
\subsubsection{Creating a query}\label{creating-a-query}}

To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species names.
This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group. Users can change this behavior by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.

\hypertarget{searching-the-database-1}{%
\subsubsection{Searching the database}\label{searching-the-database-1}}

Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).

\hypertarget{summarizing-search-results-2}{%
\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}

The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).

@@

\hypertarget{cross-validation-test}{%
\section{Cross-validation test}\label{cross-validation-test}}

We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\hypertarget{age-variation-in-source-chronograms}{%
\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
In the Fringillidae example shown here, the chronograms from Burns et al. (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).

@@

\hypertarget{primary-vs-secondary-calibrations}{%
\subsection{Primary vs Secondary calibrations}\label{primary-vs-secondary-calibrations}}

@@

\hypertarget{effects-of-phylogenetic-sampling-on-downstream-analyses}{%
\subsection{Effects of phylogenetic sampling on downstream analyses}\label{effects-of-phylogenetic-sampling-on-downstream-analyses}}

Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al. (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data. These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).

>>>>

Word count: 6111

@@

\begin{center}
\emph{Searching a Chronogram Database}
\end{center}

At the time of writing of this manuscript
(Feb 21, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

\begin{center}
\emph{Choosing a Topology}
\end{center}

@@

\begin{center}
\emph{Applying Secondary Calibrations}
\end{center}

@@

\begin{center}
\emph{Dating a Tree Topology}
\end{center}

\textbf{\emph{With no branch lengths.--}}
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008), which has proven useful for ecological analyses that require age data (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.

@@

\textbf{\emph{Dating a tree with branch lengths.--}}
Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain markers for input taxa.
By default, mined genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015). Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older nodes (Forest et al., 2005). The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
The BOLD workflow is currently only available on DateLife's R package and not on the web application.

@@

\begin{center}
\emph{Visualizing Results}
\end{center}

@@

\begin{center}
\emph{A Small Example}
\end{center}

\textbf{\emph{Creating a search query.--}}
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.

\textbf{\emph{Searching the database.--}}
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

\textbf{\emph{Summarizing search results.--}}
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).

@@

\begin{center}
\emph{An Example with the Family of True Finches}
\end{center}

\textbf{\emph{Creating a query.--}}
To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species names.
This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group. Users can change this behavior by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.

\textbf{\emph{Searching the database.--}}
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).

\textbf{\emph{Summarizing search results.--}}
The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}). Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).

@@

\begin{center}
\textsc{Cross-Validation Test}
\end{center}

We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology. Average node ages per node were applied as secondary calibrations, smoothed with the BLADJ algorithm.
We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips. Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.

\begin{center}
\textsc{Discussion}
\end{center}

@@

\begin{center}
\emph{Age Variation in Source Chronograms}
\end{center}

Conflict in estimated ages among alternative studies is common in the literature. See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
In the Fringillidae example above, the chronograms from Burns et al. (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs. \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs. S1, S5).

@@

\begin{center}
\emph{Primary vs Secondary Calibrations}
\end{center}

@@

\begin{center}
\emph{Effects of Phylogenetic Sampling on Downstream Analyses}
\end{center}

Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al. (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data. These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
